### 2015-02-12 paper/intro.tex
removed old commented out text
rewrote borrowed references introduction

%\section*{Introduction}
%Rust claims to ensure memory safety without a garbage collector.
%We would like to verify this claim.

%Rust ensures safety by combining affine types with checked borrowed references.
%Affine types ensure that every piece of memory has a unqiue owner responsible for deallocation.
%Borrowed references allow limited aliasing that avoids 
%aliased, mutable memory and dangling references
%Rust's usage of affine types for memory management is standard except for one addition.
%Rust allows consumed affine cells that are still in scope to be re-initialized and used again.
%Rust's borrowed references utilize a region system very similar to Cyclone's.
%The primary difference is in the details of the borrow checker.
%Our main goal is to prove that these differences do not introduce unsoundness.

%\section*{Patina}
%We work with a small core model of Rust called Patina, which focuses
%on the memory management aspects of Rust.
%Unlike Rust, Patina uses explicit heap deallocation and does not support closures.
%Patina has three layers: L-values, R-values, and statements.
%We will discuss them in order.

@@

\subsection*{Borrowed References}
Borrowed references are a means to provide safe, temporary aliases in Rust.
They do so by enforcing two invariants.
The first is that references are always valid.
%They do so by enforcing two invariants: 
%references are always valid and
%no memory is ever simultaneously aliased and mutable.
%As with unique pointers, these are static checks.
\begin{verbatim}
{
  let x: int; // An uninitialized stack integer
  let y: &int = &x; // ERROR! *y would point to uninitialized memory
}
\end{verbatim}
This also extends to cases where the referent falls out of scope too soon.
\begin{verbatim}
{
  let y: &int;
  {
    let x: int = 3;
    y = &x; // ERROR! x does not live long enough
  } // x would be freed here
  // *y would point to unallocated memory here
}
\end{verbatim}
The second invariant is to prevent any memory from being simultaneously aliased and mutable.
This is to avoid situtations like the following:
\begin{verbatim}
{
  let x: Option<int> = Some(3);
  let y: &int = match x {
    Some(ref z) => z,
    None => fail!(), // Impossible
  }
  // If x is mutable, then the payload 3 is both aliased and mutable
  x = None // The payload is now uninitialized so *y is a dangling pointer
}
\end{verbatim}
Therefore immutable borrows freeze the underlying data.
It cannot be mutated by anything while the borrow exists.
However, it is freely copyable.
\begin{verbatim}
{
  let x: int = 3;
  let y: &int = &x;
  let z: &int = y; // Both y and z are readable
  x = 2; // ERROR! x is frozen by y
}
\end{verbatim}
Conversely, mutable borrows forbid all other access to the underlying data.
It does allow the data to be mutated, but it is an affine datatype like unique pointers.
\begin{verbatim}
{
  let x: int = 3;
  let y: &mut int = &mut x;
  //let z: int = x; // ERROR! cannot use x while mutable borrowed
  let z: int = *y; // OK

  let w: &mut int = y; // Moves y into w. y is not uninitialized
  z = *y; // ERROR! y is uninitialized
}
\end{verbatim}
As with unique pointers, these checks are all static.

>>>>

@@

\subsection*{Borrowed Referenced REDUX}
Often, particularly in functions, we want to use some memory, but do not intend to free it.
That is, we want to borrow access to memory owned by something else.
This is fine for copyable values, like integers, 
but is tedious for noncopyable values such as unique pointers.
To avoid freeing a unique pointer argument at the end of the function body,
it must be returned from the function.
Effectively, the programmer must thread noncopyable values through their function calls.
\begin{verbatim}
fn foo(x: ~int) -> (int, ~int) {
  return (*x + 1, x)
}

{
  let x1: ~int = ~1;
  let (y,x2): (int, ~int) = foo(x);
  // y = 2, x1 is no longer valid, x2 = ~1
}
\end{verbatim}

Borrowed references allow programmers to have temporary access to a value,
but they do not confer ownership so this tedium is not necessary.
\begin{verbatim}
fn foo(x: &int) -> int {
  return (*x + 1)
}

{
  let x: ~int = ~1;
  let y: int = foo(&*x);
  // x = ~1, y = 2
}
\end{verbatim}

However, unrestricted borrowed references would open a hole in our memory safety
by enabling aliased, mutable memory.
\begin{verbatim}
{
  let x: ~int = ~1;
  let y: &~int = &x; // x and *y are the same memory location
  let z: ~int = x; // x is no longer usable - effectively x is uninitialized memory
  let w: int = **y; // ERROR! *y is uninitialized now, so **y dereferences a null pointer!
}

{
  let x: &int;
  {
    let y: int = 3;
    y = &x;
  } // x would be freed here
  let z: int = *x // ERROR *y would point to unallocated memory
}

{
  let x: Option<int> = Some(3);
  let y: &int = match x {
    Some(ref z) => z,
    None => fail!(), // Impossible
  }
  x = None // payload is now uninitialized
  let z: int = *y; // ERROR! *y points to uninitialized memory
}
\end{verbatim}

Aliased, mutable memory can be avoided in two ways: forbidding mutability or forbidding aliasing.
Correspondingly, Rust has two kinds of borrowed references.
Immutable borrows, denoted \texttt{\&}, are aliasable,
but require all memory reachable through them to be immutable for the duration of the borrow.
\begin{verbatim}
{
  let x: ~int = ~1;
  let y: &~int = &x; // this immutable borrow prevents x from being changed
  let z: ~int = x; // Rust will not let this typecheck because it would change x 
}
\end{verbatim}

Conversely, mutable borrows, denoted \texttt{\&mut}, are mutable,
but require all memory reachable through them to be \emph{only} accessible through the borrow
for the duration. That is, they guarantee unique access.
\begin{verbatim}
{
  let x: ~int = ~1;
  let y: &mut ~int = &mut x; // this mutable borrow prevents others from using x
  let z: ~int = x; // Rust will not let this typecheck because it would access x
}
\end{verbatim}

Finally, both kinds of borrowed reference require that the referent outlive the reference,
which prevents dangling references.
\begin{verbatim}
fn foo() -> &int {
  let x: int = 1;
  return &x; // Rust will not let this typecheck because it would be a dangling pointer
}
\end{verbatim}

%\subsection*{Borrowed References}
%Borrowed references are a means to provide safe, temporary aliases in Rust.
%They do so by enforcing two invariants.
%The first is that references are always valid.
%\begin{verbatim}
%{
%  let x: int; // An uninitialized stack integer
%  let y: &int = &x; // ERROR! *y would point to uninitialized memory
%}
%\end{verbatim}
%This also extends to cases where the referent falls out of scope too soon.
%\begin{verbatim}
%{
%  let y: &int;
%  {
%    let x: int = 3;
%    y = &x; // ERROR! x does not live long enough
%  } // x would be freed here
%  // *y would point to unallocated memory here
%}
%\end{verbatim}
%The second invariant is to prevent any memory from being simultaneously aliased and mutable.
%This is to avoid situtations like the following:
%\begin{verbatim}
%{
%  let x: Option<int> = Some(3);
%  let y: &int = match x {
%    Some(ref z) => z,
%    None => fail!(), // Impossible
%  }
%  // If x is mutable, then the payload 3 is both aliased and mutable
%  x = None // The payload is now uninitialized so *y is a dangling pointer
%}
%\end{verbatim}
%Therefore immutable borrows freeze the underlying data.
%It cannot be mutated by anything while the borrow exists.
%However, it is freely copyable.
%\begin{verbatim}
%{
%  let x: int = 3;
%  let y: &int = &x;
%  let z: &int = y; // Both y and z are readable
%  x = 2; // ERROR! x is frozen by y
%}
%\end{verbatim}
%Conversely, mutable borrows forbid all other access to the underlying data.
%It does allow the data to be mutated, but it is an affine datatype like unique pointers.
%\begin{verbatim}
%{
%  let x: int = 3;
%  let y: &mut int = &mut x;
%  //let z: int = x; // ERROR! cannot use x while mutable borrowed
%  let z: int = *y; // OK
%
%  let w: &mut int = y; // Moves y into w. y is not uninitialized
%  z = *y; // ERROR! y is uninitialized
%}
%\end{verbatim}
%As with unique pointers, these checks are all static.
